Binning Strategies and Related Loss for Large Data

### 2015-02-24 Paper.Rnw
couple of edits

@@

\section{Binning Data}
We will only consider binning in up to two dimensions, $X$ and $Y$. The algorithms we discuss are immediately applicable to higher dimension, but we do not feel that the paper would benefit from a more general discussion.
Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.

Binning is a two-step procedure: we first assign each data pairs $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.

 We will proceed with rectangular bins for simplicity, but other binning schemes, such as those which utilize hexagonal bins are also common \citep{scatterplots}. Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two dimensions.

\subsection{Standard Rectangular Binning Algorithm}
We will first define a binning algorithm that results in observed values being allocated to equal sized bins in a deterministic manner. We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$.  To define the set of bin centers we need only specify the center of the first bin, $x^\ast_1$ and the binwidths, $\omega_X$.  Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e. the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
% \begin{center}
% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
% \vspace{.03}
% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
% \end{center}
\begin{eqnarray}\label{standbin}
b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
\end{eqnarray}

where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.  Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.

A special case of standard binning can be used to divide the data range into $N_X$ equal sized bins.  To do so we use the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively. The first bin would then be centered at $x_{(1)} + \omega_X/2$.

>>>>

\hh{
\begin{itemize}
\item Additional literature: binning papers: standard binning, e.g.: Karl Pearson (1895) ``Contributions to the Mathematical Theory of Evolution. II. Skew Variation in Homogeneous Material"\\
David W. Scott (1979). "On optimal and data-based histograms". Biometrika 66 (3): 605–610

quantile binning: there's a lot of implementations (e.g. cut2 in package Hmisc, the rattle package, the package extracut. It would be interesting to check, if they do the same things (probably not) and mention that there is disagreement in the implementations).

As a reference I would check in 
Scott, D. W. 1992.
Multivariate Density Estimation: Theory, Practice, and Visualization. New York: Wiley.

\item make sure to verify that the binning algorithms agree with the literature (and check that the code is doing the same thing as described in the paper).
You can include a bit of discussion on the problem that histograms are over-parameterized - it's a known problem, but I can't find any paper on it at the moment.

\end{itemize}
}

@@

\section{Binning \hh{Algorithms}}
\hh{XXX Motivation first - starting with the limitiations too  negative.}

We will only consider binning in up to two dimensions, $X$ and $Y$. The algorithms we discuss are immediately applicable to higher dimension, but we do not feel that the paper would benefit from a more general discussion.
Binning in dimensions $X$ and $Y$ provides us with a more condensed form of the data that ideally preserves both the joint distribution as well as the margins, while reducing the amount of information to a fraction of the original.

Binning is a two-step procedure: we first assign each \hh{observation} $(x, y)$ to a bin center $(x^\ast,y^\ast)$, and in a second step we find the frequency of each bin center \hh{``frequency of each bin center" ... that needs a few more words, because there is only one bin center. But what you mean is that each observation assigned to the bin center now counts as one value for the bin center. XXX The next sentence explains that, so combine the two sentences a bit more}, resulting in  triples  of the form $(x^\ast, y^\ast, c)$, where $c$ is the number of all observations assigned to bin center $(x^\ast,y^\ast)$.

 We will proceed with rectangular bins for simplicity, but other binning schemes, such as \sout{those which utilize} hexagonal bins are also common \citep{scatterplots} \hh{XXX don't put citations at the end of the sentence, put them right after the most relevant phrase. in this case 'hexagonal binning'.}. Rectangular bins are advantageous because bins in $x$ and $y$ are orthogonal to each other, thus, we can present the one-dimensional case which will easily generalize to two \hh{or more} dimensions.

\hh{There are a lot of commonalities between the different univariate binning algorithms - it would make sense to write a single, more general section, and give a list of specific parameterisations for the individual binning approaches. }

\subsection{Standard Rectangular Binning Algorithm}
\hh{XXX Speed this discussion up a bit - you don't need to lay out what you will do in this section, just do it.}

\hh{Define what $X$ is, what is $x_i$}

\hh{Use only $1 \le j \le N_X$  or $j \in \{1, ..., N_X\}$. }

We \sout{will} first define a binning algorithm that \hh{allocates}  observed values to equal sized bins in a deterministic manner.
We then define a set of bin centers $\{ x_j^\ast \}$, with $1 \le j \le N_X$, where $N_X$ is the overall number of bins in dimension $X$.  To define the set of bin centers we need only specify the center of the first bin, $x^\ast_1$ and the binwidths, $\omega_X$. 
\hh{I am still not sure how bin centers are defined in the end. XXX}

\hh{why does it say binwidths? Do you want to imply that the bins have potentially different widths? Clarify that!}

Mathematically, binning corresponds to a function $b_X(.)$ that assigns to a value $x_i$ the closest bin center $x^\ast_j$ for some bin $j \in \{1, ..., N_X\}$, i.e. the binning function   $b_X(.) : x_i \rightarrow x^\ast_j$ is defined as 
% \begin{center}
% $x^\ast_i = b_X(x_i) = x_{(1)}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{(1)})/\omega_x \right\rfloor$ \hspace{.5cm} $\forall x_i < x_{(n)}$, and \\ 
% \vspace{.03}
% $x^\ast_i = b_X(x_i) = x_{(n)} - \omega_x/2$ \hspace{.5cm} $\forall x_i = x_{(n)}$
% \end{center}
\begin{eqnarray}\label{standbin}
b_X(x_i) =  x_{1}^\ast + \omega_X \cdot \left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor
\end{eqnarray}

where $x_1^\ast$ is the center of the first bin and $\lfloor . \rfloor$ denotes the floor function, $\lfloor x \rfloor$ is the largest integer value smaller than $x$. 
\hh{something went wrong with the next sentence}
An intuitive understanding of the formula is developed by noting $\left\lfloor (x_i-x_{1}^\ast)/\omega_x + 1/2 \right\rfloor$ represents the number of bin boundaries lie between $x_{1}^\ast$ and $x_i$.  Thus, the binning function is finding the location of the bin center for $x_i$ by adding the distance between $x_{1}^\ast$ and the center for $x_i$ to the known first bin center $x_{1}^\ast$.

\hh{get rid of the woulds, shoulds, cans, ...}
A special case of standard binning can be used to divide the data range into $N_X$ equal sized bins.  To do so we use the bin width, $\omega_X = (x_{(n)} - x_{(1)})/N_X$, where $x_{(1)}$ and $x_{(n)}$ denote the observed minimum and the maximum of $X$, respectively. The first bin would then be centered at $x_{(1)} + \omega_X/2$.
